Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
more

\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{def}}
%\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\sortint}{\ty{Int}}

@@

@@

The results are shown in Figures \ref{fig:sat} and \ref{fig:unsat}.
The results are divided into benchmarks triggering ``unsatisfiable'' and ``satisfiable''
responses and further into benchmarks before and after the translation $\conv$.
\jb{rephrase}
%
The raw evaluation data reveals no cases in which a solver answered
``unsatisfiable'' on a benchmark $\varphi$ and ``satisfiable'' on its
corresponding benchmark $\conv( \varphi )$, or vice versa, thereby
corroborating Theorem~\ref{thm:equi}.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command,
from the SMT-LIB 2.5 draft standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
\begin{verbatim}
(define-fun-rec s ((x Int)) Int (ite (<= x 0) 0 (+ x (s (- x 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\, 0,\, \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0, \con{s}( x ) \teq 0, \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using the command line parameter ``--fmf-fun"),
then it will replace its list of known assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it used for recursive function definitions,
that is, it will output the following model.
\ajr{prefer having the full output like this here?}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.

{\small
\begin{verbatim}
(model
(define-fun s (($x1 Int)) Int 
               (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
               (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
               (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
               (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
               (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
(define-fun c () Int 14)
)
\end{verbatim}
}

Notice the option ``--fmf-fun" effectively instructs the SMT solver to assume that the functions provided by the user are admissible.
If this option is used for a set of functions definitions that is \emph{not} admissible, then the solver is not model-sound, that is,
it may return ``satisfiable" for unsatisfiable inputs.
As such,
this option targets users who wish to determine the satisfiability of inputs that involving function definitions that are known to be consistent,
whereas it does not target users who wish to check the consistency of a set of function definitions.

>>>>

\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{dfn}}
%\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\sortint}{\ty{Int}}

@@

\begin{verbatim}
this option targets users who wish to determine the satisfiability of inputs
that involving function definitions that are known to be consistent, whereas it
does not target users who wish to check the consistency of a set of function
definitions.
\end{verbatim}

@@

The results are shown in Figures \ref{fig:sat} and \ref{fig:unsat}.
The results are divided into benchmarks triggering ``unsatisfiable'' and ``satisfiable''
responses and further into benchmarks before and after the translation $\conv$.
The raw evaluation data reveals no cases in which a solver answered
``unsatisfiable'' on a benchmark $\varphi$ and ``satisfiable'' on its
corresponding benchmark $\conv( \varphi )$, or vice versa, thereby
corroborating Theorem~\ref{thm:equi}.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command,
from the SMT-LIB 2.5 draft standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
\begin{verbatim}
(define-fun-rec s ((n Int)) Int (ite (<= n 0) 0 (+ n (s (- n 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. For our example, it will output the following
model:
\ajr{prefer having the full output like this here?}
\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}
%
The \texttt{--fmf-fun} option instructs CVC4 to assume that the
functions provided by the user are admissible.
If this condition is not met, the solver may answer ``satisfiable'' on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, CVC4 will still find the same ``model''
if \texttt{--fmf-fun} is specified.
